---
title: "Manual"
author:
- N. Frerebeau
- B. Lebrun
- G. Guérin
- C. Lahaye
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Manual}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(gamma)
library(magrittr)
```

## Overview

`gamma` is intended to process in-situ gamma-ray spectrometry measurements for luminescence dating. This package allows to import, inspect and (automatically) correct the energy scale of the spectrum. It provides methods for estimating the gamma dose rate by the use of a calibration curve. This package only supports Canberra CNF and TKA files.

Typical workflow:

1. Import spectrum data with `read()`
2. Inspect spectrum with `plot()`
3. Calibrate the energy scale with `calibrate()`
4. Estimate the gamma dose rate with `predict()`

The estimation of the gamma dose rate requires a calibration curve. This package contains several built-in curves, but as these are instrument-specific you may need to build your own (see `help(fit)`). Check out the following vignettes for an overview of the fitting process.

```{r calib-dose, eval=FALSE}
# IRAMAT-CRP2A LaBr (#1)
utils::vignette("CRP2A#1", package = "gamma")

# CEREGE NaI (#1)
utils::vignette("CEREGE#1", package = "gamma")
```

## Import file
```{r import, fig.width=7, fig.height=5, fig.align="center"}
# Automatically skip the first chanels
# Import a CNF file
cnf_test <- system.file("extdata/test_CNF.cnf", package = "gamma")
(spc_cnf <- read(cnf_test, skip = TRUE))

# Import a TKA file
tka_test <- system.file("extdata/test_TKA.tka", package = "gamma")
(spc_tka <- read(tka_test, skip = TRUE))

# Import all files in a given directory
files_test <- system.file("extdata/", package = "gamma")
(spc <- read(files_test, skip = TRUE))
```

## Inspect spectrum
```{r inspect, fig.width=7, fig.height=5, fig.align="center"}
# Plot TKA spectrum
plot(spc_cnf) +
  ggplot2::theme_bw()

# Plot spectra
plot(spc, facet = TRUE) +
  ggplot2::theme_bw()
```

## Calibrate the energy scale
### Estimate and remove baseline
SNIP algorithm [@ryan1988; @morhac1997; @morhac2008]

```{r baseline-estimate, fig.width=7, fig.height=5, fig.align="center"}
# Estimate baseline of a single file
baseline <- estimateBaseline(spc_tka)

# Plot spectrum + baseline
plot(spc_tka, baseline) +
  ggplot2::theme_bw()
```

```{r baseline-remove, fig.width=7, fig.height=5, fig.align="center"}
# Substract the previously estimated baseline
spc_clean1 <- spc_tka - baseline
# Or, remove the baseline in one go
spc_clean2 <- removeBaseline(spc_tka)

identical(spc_clean1, spc_clean2)

# Plot spectrum
plot(spc_clean1) +
  ggplot2::theme_bw()
```

```{r baseline-remove-multi, eval=FALSE}
# Remove the baseline of multiple spectra in one go
spc_clean <- removeBaseline(spc)
```

### Automatic peak detection
```{r peak-detection, fig.width=7, fig.height=5, fig.align="center", fig.show="hold"}
# Detect peaks in a single file
peak_index <- findPeaks(spc_tka)

# Plot spectrum
plot(peak_index) +
  ggplot2::theme_bw()
```

```{r peak-detection-table, echo=FALSE}
# Get peak positions
knitr::kable(
  peak_index[["peaks"]],
  digits = 3,
  col.names = c("Chanel", "Energy [keV]", "Count", "Count rate [1/s]"),
  caption = "Automatic peak detection"
)
```

### Peak parameters estimation
```{r peak-fit-auto, fig.width=7, fig.height=5, fig.align="center"}
# Fit peaks found by automatic detection
peak_fit_auto <- fitPeaks(peak_index)

plot(peak_fit_auto) +
  ggplot2::labs(title = "Fitted peaks from automatic detection") +
  ggplot2::theme_bw()
```

```{r peak-fit-auto-table, echo=FALSE}
# Get peak parameters
knitr::kable(
  peak_fit_auto[["coefficients"]],
  digits = 3,
  col.names = c("Mean (chanel)", "Std. dev. (chanel)", "Height (count)"),
  caption = "Peak parameters estimate"
)
```

```{r peak-fit-manual, fig.width=7, fig.height=5, fig.align="center"}
# Fit peaks at a given chanel
peak_fit_chanel <- fitPeaks(spc_tka, peaks = c(86, 496, 876))

## Plot results
plot(peak_fit_chanel) +
  ggplot2::labs(title = "Fitted peaks from fixed starting positions") +
  ggplot2::theme_bw()
```

```{r peak-fit-manual-table, echo=FALSE}
## Get peak positions
knitr::kable(
  peak_fit_chanel[["coefficients"]],
  digits = 3,
  col.names = c("Mean (chanel)", "Std. dev. (chanel)", "Height (count)"),
  caption = "Peak positions"
)
```

### Energy scale calibration
```{r calibrate-fixed, fig.width=7, fig.height=5, fig.align="center"}
calib_lines <- list(
  Pb = c(chanel = 86, energy = 238),
  K40 = c(chanel = 496, energy = 1461),
  Cs = c(chanel = 876, energy = 2614.5)
)

# Calibrated spectrum using fixed lines
tka_scaled1 <- calibrate(spc_tka, lines = calib_lines)

# Calibrated spectrum using peak models
tka_scaled2 <- calibrate(peak_fit_chanel, lines = c(238, 1461, 2614.5))

plot(tka_scaled1, tka_scaled2) +
  ggplot2::labs(title = "Calibrated spectra") +
  ggplot2::theme_bw()
```


## Gamma dose rate estimation
```{r dose-rate, fig.width=7, fig.height=5, fig.align="center"}
# Load the calibration curve
data(BDX1)

# Estimate the gamma dose rate
dose_rate <- predict(BDX1, tka_scaled2, simplify = TRUE)
```

```{r dose-rate-table, echo=FALSE}
knitr::kable(
  dose_rate,
  digits = 3,
  col.names = c("Dose", "Dose error"),
  caption = "Gamma dose rate estimates"
)
```

## References
